Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-05 paper.tex
related work

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence.  Recent work in program induction show
that deep networks can be very successful at learning to write certain
kinds of programs (e.g., RobustFill:~\cite{devlin2017robustfill} and
DeepCoder:~\cite{balog2016deepcoder}), as can symbolic search
techniques (e.g., Metagol:~\cite{muggleton2015meta} and 
FlashFill:~\cite{gulwani2011automating}).
However, the success of these approaches -- both symbolic and neural --
relies critically upon a hand-engineered \emph{Domain Specific Language} (DSL).
The DSL is an inventory of restricted programming primitives,
crafted by an engineer,
which imparts domain specific knowledge about the space of programs.
To what extent can we dispense with hand-engineered DSLs?

@@

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}(a): black lines):
A DSL $\mathcal{D}$, equipped with a real-valued weight vector $\theta$,
determine a distribution over programs
from which a latent program is sampled,
which finally determines the agent's observations.
Our goal will be both to infer the programs that best explain our observations,
and at the same time induce a DSL
that distills the commonalities across all of the programs that solve the tasks.
Inducing the DSL means learning the symbolic structure of a latent programming language,
and so we are confronted with two distinct structure learning problems:
the structure of the programs and the structure of the DSL.

@@

How should we perform inference for the model
in~\ref{graphicalModel}(a)?  We take the intuitive strategy of exploiting the
conditional independence of the programs conditioned on the DSL, and
so alternate between estimating the DSL ($\mathcal{D},\theta$) and
estimating a posterior distribution over $p$ by searching for
programs written using $\mathcal{D}$.  In practice searching the
space of programs is difficult, so we use the recognition model to
learn how to efficiently search for programs.

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in each of these domains (Table~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Subset of induced DSL\\\midrule
   \\\bottomrule
 \end{tabular}
   \end{table}

This makes our work in instance of
 \emph{Bayesian Program
  Learning} (BPL; see~\citep{lake2013one,Dechter:2013:BLV:2540128.2540316,ellis2016sampling,DBLP:conf/icml/LiangJK10}).
 Alongside this generative model, our algorithm learns a bottom-up recognition model (Figure~\ref{graphicalModel}(b)).
 The recognition model is a neural network $q$ that regresses from programming tasks $x$
 to a distribution over programs.

TODO: talk about existing works as special cases of, and points of inspiration for, \system:
\begin{itemize}
  \item Fixed DSL, no recognition model, learn parameters $\theta$ of an inductive bias over programs\citep{menon2013machine,singh2015predicting,learningToRank}
  \item The EC algorithm and related work \citep{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,DBLP:conf/ecai/LinDETM14} has no recognition model
    $q$;
    % because this is a primary inspiration, might be worth further mention:
    % different program representation (routing vs. substitution; simple vs polymorphic types);
    % different grammar induction (sequitur vs. fragment grammar)
  \item DeepCoder and RobustFill \citep{balog2016deepcoder,devlin2017robustfill} both fix the
    DSL $\mathcal{D}$ and does not use $\theta$ in favor of solely relying on the neural
    network $q$.
    % although RobustFill has very different enumeration
\end{itemize}

@@

Our goal in this work is to both induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs that
solve the tasks, (2) learning a better neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that each of these three steps can bootstrap off each other:

\begin{itemize}
\item Searching for programs: Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.
  \item Learning a recognition model: The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
  \item Improving the DSL: We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.
\end{itemize}

@@

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program induction problem.
It has at its disposal a \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL $\mathcal{D}$ that distills the commonalities across all of the programs that solve the tasks.

>>>>

Automatically inducing programs from examples is a long-standing goal
of artificial intelligence.  Recent work in program induction show
that deep networks can be very successful at learning to write certain
kinds of programs (e.g., RobustFill:~\cite{devlin2017robustfill} and
DeepCoder:~\cite{balog2016deepcoder}), as can symbolic search
techniques (e.g., Metagol:~\cite{muggleton2015meta} and 
FlashFill:~\cite{gulwani2011automating}).
However, the success of these approaches -- both symbolic and neural --
relies crucially upon a hand-engineered \emph{Domain Specific Language} (DSL).
The DSL is an inventory of restricted programming primitives,
that imparts domain specific knowledge about the space of programs.
To what extent can we dispense with hand-engineered DSLs?

@@

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}(a): black lines):
A DSL $\mathcal{D}$, equipped with a real-valued weight vector $\theta$,
give a distribution over programs
from which a latent program is sampled,
which then determines the agent's observations.
Our goal will be both to infer the programs that best explain our observations,
and at the same time induce a DSL
that distills the commonalities across all of the programs that solve the tasks.
Inducing the DSL means learning the symbolic structure of a latent programming language,
and so we are confronted with two distinct structure learning problems:
the structure of the programs and the structure of the DSL.

@@

How should we perform inference for the model
in~\ref{graphicalModel}(a)?  We take the intuitive strategy of exploiting the
conditional independence of the programs conditioned on the DSL, and
so alternate between estimating the DSL ($\mathcal{D},\theta$) and
estimating a posterior distribution over $p$ by searching for
programs written using $\mathcal{D}$.  In practice searching for
programs is difficult, so we use the recognition model to
learn how to search efficiently.

We apply \system to four domains:
 building Boolean circuits; symbolic regression; FlashFill-style~\cite{gulwani2011automating} string processing problems; and Lisp-style functions on lists.
 For each of these we deliberately provide an impoverished
 set of programming primitives,
 and show that our algorithm discovers
 its own domain-specific vocabulary for expressing solutions in each of these domains (Table~\ref{initialExampleDSL}).
 \begin{table}
    \begin{tabular}{ll}
   \toprule
   Domain&Part of the learned DSL\\\midrule
   Circuits&\texttt{(nand (nand a a) (nand b b))}\\
   &\hspace{0.5cm} \emph{(an OR gate)}\\
   Regression& \texttt{(+ (* real x) real)} \\
      &\hspace{0.5cm} \emph{(a linear function of x)}\\
   Strings& \texttt{(join '' (split c s))}\\
   &\hspace{0.5cm} \emph{(delete occurrences of c in string s)}\\
   Lists&
   \\\bottomrule
    \end{tabular}
    \caption{ Examples of structure found in DSLs  learned by our algorithm. \system builds a new DSL by discovering and reusing useful fragments of code.}\label{initialExampleDSL}
 \end{table}

 \subsection{Related Work}
 Our work is far from the first to learn to learn programs,
 an  idea that goes back to Solomonoff~\cite{solomonoff1989system}:

 \noindent \emph{Neural recognition models for program induction:} Much recent work in the ML community has
 focused on engineering neural networks that regress from
 input/output examples to programs~\cite{devlin2017robustfill,devlin2017neural,menon2013machine,balog2016deepcoder}. This family of work is analogous to a lesioned version of
 \system where $(\mathcal{D},\theta)$ is held fixed and $q$ is learned.

 \noindent \emph{Predicate Invention in ILP:} The Inductive Logic Programming (ILP) community
 has long worked on the problem of inventing new primitive predicates~\cite{DBLP:conf/ecai/LinDETM14,muggleton2015meta}, which is analogous to our model with $\theta$ held fixed and $q$ ablated.

 \noindent \emph{Inventing new subroutines for program induction:}
 Several program induction algorithms, most prominently the EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}, take as their goal to learn new, reusable subroutines that are shared in a multitask setting. We find this work inspiring and motivating,
 and go beyond it along two dimensions: (1) we propose more sophisticated algorithms for
 inducing reusable structure, based on a formalism called \emph{Fragment Grammars}~\cite{tim};
 and (2) we show how to combine these techniques with bottom-up neural recognition models.
 Other instances of this related idea are~\cite{DBLP:conf/icml/LiangJK10} and Schmidhuber's OOPS model~\cite{schmidhuber2004optimal}.

Our work is an instance of
 \emph{Bayesian Program
  Learning} (BPL; see~\citep{lake2013one,Dechter:2013:BLV:2540128.2540316,ellis2016sampling,DBLP:conf/icml/LiangJK10}).

%% \begin{itemize}
%%   \item Fixed DSL, no recognition model, learn parameters $\theta$ of an inductive bias over programs\citep{menon2013machine,singh2015predicting,learningToRank}
%%   \item The EC algorithm and related work \citep{Dechter:2013:BLV:2540128.2540316,DBLP:conf/icml/LiangJK10,DBLP:conf/ecai/LinDETM14} has no recognition model
%%     $q$;
%%     % because this is a primary inspiration, might be worth further mention:
%%     % different program representation (routing vs. substitution; simple vs polymorphic types);
%%     % different grammar induction (sequitur vs. fragment grammar)
%%   \item DeepCoder and RobustFill \citep{balog2016deepcoder,devlin2017robustfill} both fix the
%%     DSL $\mathcal{D}$ and does not use $\theta$ in favor of solely relying on the neural
%%     network $q$.
%%     % although RobustFill has very different enumeration
%% \end{itemize}

@@

Our goal  is to induce a DSL and find good programs solving each of the tasks.
Our strategy is to iterate through three steps: (1) searching for programs that
solve the tasks, (2) learning a better neural recognition model -- which we use
to accelerate the search over programs -- and (3) improving the DSL.
The key observation here is that each of these three steps can bootstrap off each other:

\begin{description}
\item[ Searching for programs:] Our program search  is informed by both the DSL and the recognition model.
When these improve we find better programs.
  \item[ Learning a recognition model:] The recognition model is trained both on samples from the DSL and on programs found by the search procedure. As the DSL improves and as search finds more programs, the recognition model gets both more data to train on and better data.
  \item[Improving the DSL:] We induce a DSL from the programs we have found so far which solve the tasks;
  as we solve more tasks, we can hone in on richer DSLs that more closely match the domain.
\end{description}

@@

\system takes as input a set of \emph{tasks}, written $X$, each of which is a program induction problem.
It has at its disposal a \emph{likelihood model}, written $\probability[x|p]$, which scores the likelihood of a task $x\in X$ given a program $p$.
Its goal is to solve each of the tasks by writing a program,
and also to infer a DSL $\mathcal{D}$.
